Species motif participation provides unique information about species risk of extinction

### 2023-05-12 manuscript/reorder_main_text.tex
fixed SI refs

        We generated simulated networks using the niche model, which recreates the structure of empirical networks well~\citep{Williams2000,Stouffer2007}.
        To capture a range of network architectures similar to those in well-studied empirical networks~\citep{Dunne2002,Dunne2002e}, we simulated networks with sizes (S) ranging from 50 to 100 species (in steps of 10) and connectances (C) ranging from 0.02 to 0.18 (in steps of 0.04). 
        We removed any biologically unlikely networks (i.e., those with extremely long paths between any consumer and basal resources~\citep{Borrelli2014} and replaced them with new simulated networks, repeating the process until we obtained 100 suitable networks in each combination of size and connectance.
        All networks were generated using the function ``nichemodel'' within the Julia~\citep{Bezanson2017julia} package \emph{BioEnergeticFoodWebs}~\citep{bioenergfw,Delmas2017}.     For full details and a comparison to empirical networks, see \emph{Appendix S2}.

        % To make the simulated networks compatible with the Bayesian simulation framework, all networks were rendered acyclic following~\citet{Allesina2009functional} and species topologically sorted from lowest to highest trophic level (\emph{Appendix S2}). 
        %  These processing steps allow the strictly bottom-up calculation of probability of persistence.
        We described the motif participation of each consumer species in each of these simulated networks before applying any disturbance. 
        A species' motif participation (Fig.~\ref{fig:concept}) is the vector of frequencies with which it appears in each type of motif present in a network~\citep{Stouffer2012}.
        As only four unique three-species motifs can appear in acyclic networks, these are four-dimensional vectors. 
        Here we are interested in the relative frequencies of each motif rather than the total number of motifs and therefore normalized participation vectors by dividing each count by the total number of motifs the species appears in (such that the vector for each species sums to 1).
        Note, however, that this normalization does not fully control for differences due to degree or connectance~\citep{Cirtwill2022Oikos}.

@@

        Bayesian networks also require the creation of acyclic networks so that a strict bottom-up sequence can be followed. 
        This means that cycles where species A eats species B and vice versa or where A eats B eats C eats A must be broken (see \emph{Appendix SX} for details).
        Since cyclic and acyclic versions of a network have the same properties in terms of secondary extinctions and the effects on network robustness are very limited~\citep{Allesina2009functional}, this change will not strongly affect consumers' extinction risk in our simulations.

@@

        The response of the consumer to changes in $f$ can take any form (\emph{Appendix S1}). Here we use a sigmoid non-linear function ($\alpha > 1, \beta > 1$), the type of response that most accurately captures the secondary extinctions produced by dynamical models (\citealp[]{Eklof2013}).
		The probability of extinction for each species $i$ can therefore be represented using the cumulative density function of a beta distribution:
		\begin{equation}
		P(\lnot i|f) = \pi + (1 - \pi) \frac{B(f;\alpha,\beta)}{B(\alpha,\beta)}.
				\label{betafunc}
        \end{equation}

@@

	\subsection*{Statistical analysis} 
	% Appendix 3, with numbered subheadings.

@@

            To test whether these general trends were consistent across networks, we visually examined the distribution of slopes of simplified regressions fit to a single network and level of disturbance.
            These models included only participation in the focal motif ($\rho_{i}$ above) and again used a binomial error distrubution.
            To aid in interpretation of these single-network models with respect to network size and connectance, we fit an additional set of GLMs including proportion of motif, strength of disturbance, network size, and connectance across all networks (\emph{Appendix S6}).

@@

            We also confirmed previously-known relationships between persistence and other measures of network structure or a species' participation in the network.
            To identify the relationship between large-scale network structure and mean persistence within a network, and how the strength of disturbance might affect this relationship, we fit a GLM relating mean persistence to network size, connectance, disturbance, and their interactions.
            As the three-way interaction was significant, we did not simplify the model. 
            To confirm known relationships between persistence and in-degree or trophic level in our simulated networks and identify how they might vary with different strengths of disturbance, we fit two GLMs with binomial error distribution relating a consumer's persistence to in-degree or STL, disturbance, and the interaction between them (\emph{Appendix S3}, Eq. S2-3).
            As with the GLMs relating persistence to motif participation and motif profiles, we initially fit GLMMs including random effects of network size and connectance and network identity but removed these terms due to model singularity.

@@

            Since a network's motif profile and a species' motif participation reflect the same network as other measures (size, connectance, degree, and trophic level), these measures are not independent. 
            We identified how the frequencies of motifs in the network's motif profile were related to network size, connectance, and their interaction using a set of four GLMs with binomial error distribution (Appendix \emph{S3}, eq. S4).
            We identified how the frequencies of motifs in a consumer's motif participation vector were related to network size, connectance, and their interaction using a similar set of four GLMs with binomial error distribution.
            Finally, we fit two additional sets of GLMs (eight total) relating participation in each motif to the consumer's in-degree or trophic level (Appendix \emph{S3}, Eq. S5-6).

@@

        A species' probability of persistence (hereafter `persistence') was significantly related to its motif participation (Table S1, \emph{Appendix S4}).
        Across all levels of disturbance, species were more likely to persist if they participated more frequently in the direct competition motif and less frequently in the omnivory motif (Fig.~\ref{fig:prop_lmer_all}).
        The relationships between persistence and participation in the other two motifs varied with the strength of disturbance.
        At low levels of disturbance, species were more likely to persist if they participated more frequently in the three-species chain motif and less frequently in the apparent competition motif.
        These trends were reversed at high levels of disturbance.

        Within these overall trends, there was substantial variation in the relationships between persistence and motif participation between networks with different overall network structures (Fig.~\ref{fig:density_prop}).
        Networks with high and low connectances had different distributions of persistence-motif participation relationships, while network size had a much smaller effect (\emph{Appendix SX}).
        The relationship between participation in the omnivory motif and persistence was especially variable.
        Higher proportions of omnivory were associated with lower probabilities of persistence when connectance was low, regardless of network size or strength of disturbance.
        At moderate connectance, however, higher proportions of omnivory were associated with greater probabilities of persistence when disturbance was weak.
        In the most-connected networks, higher proportions of omnivory were associated with greater probabilities of persistence at all levels of disturbance.

@@

        Consumers with higher trophic level (STL) were less likely to persist (Fig.~\ref{fig:lm_CS}, Table S3, \emph{Appendix S6}).
        The relationship between in-degree and persistence depended on the strength of disturbance: higher in-degree was associated with greater persistence when basal species had low probability of extinction but lower persistence when basal species were highly likely to go extinct.

        The average persistence of all consumers was lower in networks with higher connectance, did not vary significantly with network size (Table S6, \emph{Appendix S10}), and  varied with the frequencies of different motifs in this profile.
        Specifically, average consumer persistence was higher in networks with higher proportions of the apparent competition motif, lower in networks with higher proportions of the omnivory motif, and was not significantly related to the frequencies of the three-species chain or direct competition motifs (Fig.~\ref{fig:lm_CS}, Table S4, \emph{Appendix S8}). 
        There was no significant interaction between the proportion of any motif in the network profile and level of disturbance (Table S4), \emph{Appendix S8}).

@@

        Omnivory motif participation generally increased with increasing connectance and network size, but decreased with increasing network size in the most-connected networks (Fig. S2 and Table S2, \emph{Appendix S5}).
        Participation in the apparent competition and three-species chain motifs decreased strongly with increasing connectance but showed little change with changing network size.
        Direct competition motif participation decreased with increasing connectance in small networks and increased with increasing network size in highly-connected networks, but showed little change with connectance in large networks or with network size in less-connected networks.

>>>>

        We generated simulated networks using the niche model, which recreates the structure of empirical networks well~\citep{Williams2000,Stouffer2007}.
        To capture a range of network architectures similar to those in well-studied empirical networks~\citep{Dunne2002,Dunne2002e}, we simulated networks with sizes (S) ranging from 50 to 100 species (in steps of 10) and connectances (C) ranging from 0.02 to 0.18 (in steps of 0.04). 
        We removed any biologically unlikely networks (i.e., those with extremely long paths between any consumer and basal resources~\citep{Borrelli2014} and replaced them with new simulated networks, repeating the process until we obtained 100 suitable networks in each combination of size and connectance.
        All networks were generated using the function ``nichemodel'' within the Julia~\citep{Bezanson2017julia} package \emph{BioEnergeticFoodWebs}~\citep{bioenergfw,Delmas2017}.     For full details and a comparison to empirical networks, see \emph{Appendix S1}.

        We described the motif participation of each consumer species in each of these simulated networks before applying any disturbance. 
        A species' motif participation (Fig.~\ref{fig:concept}) is the vector of frequencies with which it appears in each type of motif present in a network~\citep{Stouffer2012}.
        As only four unique three-species motifs can appear in acyclic networks, these are four-dimensional vectors. 
        Here we are interested in the relative frequencies of each motif rather than the total number of motifs and therefore normalized participation vectors by dividing each count by the total number of motifs the species appears in (such that the vector for each species sums to 1).
        Note, however, that this normalization does not fully control for differences due to degree or connectance~\citep{Cirtwill2022Oikos}.

@@

        Bayesian networks also require the creation of acyclic networks so that a strict bottom-up sequence can be followed. 
        This means that cycles where species A eats species B and vice versa or where A eats B eats C eats A must be broken (see \emph{Appendix S1.3} for details).
        Since cyclic and acyclic versions of a network have the same properties in terms of secondary extinctions and the effects on network robustness are very limited~\citep{Allesina2009functional}, this change will not strongly affect consumers' extinction risk in our simulations.

@@

        The response of the consumer to changes in $f$ can take any form (\emph{Appendix S2.1}). Here we use a sigmoid non-linear function ($\alpha > 1, \beta > 1$), the type of response that most accurately captures the secondary extinctions produced by dynamical models (\citealp[]{Eklof2013}).
		The probability of extinction for each species $i$ can therefore be represented using the cumulative density function of a beta distribution:
		\begin{equation}
		P(\lnot i|f) = \pi + (1 - \pi) \frac{B(f;\alpha,\beta)}{B(\alpha,\beta)}.
				\label{betafunc}
        \end{equation}

@@

	\subsection*{Statistical analysis}

@@

            To test whether these general trends were consistent across networks, we visually examined the distribution of slopes of simplified regressions fit to a single network and level of disturbance.
            These models included only participation in the focal motif ($\rho_{i}$ above) and again used a binomial error distrubution.
            To aid in interpretation of these single-network models with respect to network size and connectance, we fit an additional set of GLMs including proportion of motif, strength of disturbance, network size, and connectance across all networks (\emph{Appendix S4.1}).

@@

            We also confirmed previously-known relationships between persistence and other measures of network structure or a species' participation in the network.
            To identify the relationship between large-scale network structure and mean persistence within a network, and how the strength of disturbance might affect this relationship, we fit a GLM relating mean persistence to network size, connectance, disturbance, and their interactions.
            As the three-way interaction was significant, we did not simplify the model. 
            To confirm known relationships between persistence and in-degree or trophic level in our simulated networks and identify how they might vary with different strengths of disturbance, we fit two GLMs with binomial error distribution relating a consumer's persistence to in-degree or STL, disturbance, and the interaction between them (\emph{Appendix S3.1}, Eq. S2-3).
            As with the GLMs relating persistence to motif participation and motif profiles, we initially fit GLMMs including random effects of network size and connectance and network identity but removed these terms due to model singularity.

@@

            Since a network's motif profile and a species' motif participation reflect the same network as other measures (size, connectance, degree, and trophic level), these measures are not independent. 
            We identified how the frequencies of motifs in the network's motif profile were related to network size, connectance, and their interaction using a set of four GLMs with binomial error distribution (Appendix \emph{S3.2}, eq. S4).
            We identified how the frequencies of motifs in a consumer's motif participation vector were related to network size, connectance, and their interaction using a similar set of four GLMs with binomial error distribution.
            Finally, we fit two additional sets of GLMs (eight total) relating participation in each motif to the consumer's in-degree or trophic level (Appendix \emph{S3.2}, Eq. S5-6).

@@

        A species' probability of persistence (hereafter `persistence') was significantly related to its motif participation (Table S1, \emph{Appendix S5}).
        Across all levels of disturbance, species were more likely to persist if they participated more frequently in the direct competition motif and less frequently in the omnivory motif (Fig.~\ref{fig:prop_lmer_all}).
        The relationships between persistence and participation in the other two motifs varied with the strength of disturbance.
        At low levels of disturbance, species were more likely to persist if they participated more frequently in the three-species chain motif and less frequently in the apparent competition motif.
        These trends were reversed at high levels of disturbance.

        Within these overall trends, there was substantial variation in the relationships between persistence and motif participation between networks with different overall network structures (Fig.~\ref{fig:density_prop}).
        Networks with high and low connectances had different distributions of persistence-motif participation relationships, while network size had a much smaller effect (\emph{Appendix S4.2}).
        The relationship between participation in the omnivory motif and persistence was especially variable.
        Higher proportions of omnivory were associated with lower probabilities of persistence when connectance was low, regardless of network size or strength of disturbance.
        At moderate connectance, however, higher proportions of omnivory were associated with greater probabilities of persistence when disturbance was weak.
        In the most-connected networks, higher proportions of omnivory were associated with greater probabilities of persistence at all levels of disturbance.

@@

        Consumers with higher trophic level (STL) were less likely to persist (Fig.~\ref{fig:lm_CS}, Table S3, \emph{Appendix S6.1}).
        The relationship between in-degree and persistence depended on the strength of disturbance: higher in-degree was associated with greater persistence when basal species had low probability of extinction but lower persistence when basal species were highly likely to go extinct.

        The average persistence of all consumers was lower in networks with higher connectance, did not vary significantly with network size (Table S4, \emph{Appendix S6.1}), and  varied with the frequencies of different motifs in this profile.
        Specifically, average consumer persistence was higher in networks with higher proportions of the apparent competition motif, lower in networks with higher proportions of the omnivory motif, and was not significantly related to the frequencies of the three-species chain or direct competition motifs (Fig.~\ref{fig:lm_CS}, Table S5, \emph{Appendix S6.2}). 
        There was no significant interaction between the proportion of any motif in the network profile and level of disturbance).

@@

        Omnivory motif participation generally increased with increasing connectance and network size, but decreased with increasing network size in the most-connected networks (Fig. S7 and Table S6, \emph{Appendix S7}).
        Participation in the apparent competition and three-species chain motifs decreased strongly with increasing connectance but showed little change with changing network size.
        Direct competition motif participation decreased with increasing connectance in small networks and increased with increasing network size in highly-connected networks, but showed little change with connectance in large networks or with network size in less-connected networks.
